---
title: " Product Grouping Analysis"
author: "Pavel Logačev"
date: "r Sys.Date()"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
engine: knitr
editor: visual
---

```{r setup, eval=FALSE, include=FALSE}
this_dir <- dirname(knitr::current_input())
setwd(this_dir)
```

### 1.1 Notebook Settings

```{python message=FALSE, warning=FALSE}
## set transformer model 
#py_model_name = 'all-MiniLM-L6-v2' # primarily English
#py_model_name = 'paraphrase-multilingual-MiniLM-L12-v2' # multilingual
py_model_name = 'LaBSE' 
#py_model_name = 'xlm-r-100langs-bert-base-nli-stsb-mean-tokens' 
py_model_name = 'intfloat/multilingual-e5-base'


# runner ups to try, according to ChatGPT:
# - XLM-RoBERTa
# - mBART-50
# - MT5 (Multilingual T5)
# - ByT5 (handles raw text at the byte level, ideal for multilingual similarity tasks).
# - mLUKE (multilingual version of LUKE, good for named entity recognition and contextual understanding).
# - mBERT (baseline multilingual BERT, effective but less performant than XLM-R).

n_cores = 8
```

### 1.2 Load Packages

```{python message=FALSE, warning=FALSE}
import os
os.environ["TOKENIZERS_PARALLELISM"] = "true"
os.environ["SIM_EMBEDDINGS_CORES"] = f"{n_cores}"

import pandas as pd
import polars as pl

import random
import pickle
import numpy as np
from tqdm import tqdm
import sentence_transformers as st

import nltk
from nltk.tokenize import word_tokenize

import gensim
from gensim.models import Word2Vec

from sklearn.cluster import MiniBatchKMeans
from sklearn.metrics.pairwise import cosine_similarity
from sklearn.decomposition import PCA

import multiprocessing
multiprocessing.set_start_method("forkserver", force=True)

import torch
torch.set_num_threads(n_cores)

from plotnine import *

from collections import defaultdict

import json
from scipy.cluster.hierarchy import linkage, to_tree


#from joblib import Memory
#compute_memory = Memory("./cache", verbose=0)


```

### 1.3 Globally Used Code

```{python message=FALSE, warning=FALSE}
# Enable the underscore notation for polars columns, similar to ibis and dplyr (i.e., `pl.col("column_name")` -> `_.column_name`)
class PolarsColumnNamespace:
    def __getattr__(self, name):
        return pl.col(name)

# Enable _ as a shorthand for that class
_ = PolarsColumnNamespace()
```



```{python message=FALSE, warning=FALSE}
import os
import pickle
from functools import wraps

def cache(func):
    """
    Decorator: Caches the result of func to a pickle file if fname_cache is given.
    The decorated function must accept fname_cache as a keyword argument.
    """
    @wraps(func)
    def wrapper(*args, **kwargs):
        fname_cache = kwargs.get("fname_cache", None)
        
        if fname_cache is not None and os.path.exists(fname_cache):
            print(f"[CACHE] Loading from {fname_cache}")
            with open(fname_cache, "rb") as f:
                return pickle.load(f)
        
        print(f"[CACHE] Computing and caching to {fname_cache}")
        result = func(*args, **kwargs)
        
        if fname_cache is not None:
            with open(fname_cache, "wb") as f:
                pickle.dump(result, f)
            print(f"[CACHE] Saved to {fname_cache}")
        
        return result

    return wrapper
```





## 2. Prepare Data

### 2.1 Read Transactional Data

mention that these are two datasets from UCI

```{python message=FALSE, warning=FALSE}
df1 = pl.read_excel('./data/online+retail/Online Retail.xlsx')
df2 = pl.read_excel('./data/online+retail+ii/online_retail_II.xlsx')
df2 = df2.rename({"Invoice": "InvoiceNo","Price": "UnitPrice", "Customer ID": "CustomerID"})
df = df1.vstack(df2)
```

### 2.2 Clean Up Dataset

#### 2.2.1 Missing Invoice IDs
- A number of records do not have an invoice id. We'll exclude them, because we cannot use them to determine the basket co-occurrence frequencies for the respective entries.
- This applies to $19500$ rows, i.e., $1.8\%$ of the records

```{python message=FALSE, warning=FALSE}
nrows_null = df.filter( _.InvoiceNo.is_null() ).shape[0]
nrows_total = df.shape[0]
(nrows_null, nrows_null / nrows_total)
```

```{python message=FALSE, warning=FALSE}
# drop rows with null InvoiceNo
df = df.filter( ~_.InvoiceNo.is_null() )
```

#### 2.2.2 Missing or Incorrect Product Descriptions
- Some product descriptions indicate adjustments or comments rather than product names. In some cases, additional information is provided, and in others, they have been initially mislabeled.
- We'll exclude the corresponding transactions from the data.
- Exploratory analysis showed that nearly all proper product descriptions are uppercase, and those that are not fully uppercase are in title case.
    - Thus, one heuristic to identify anomalous descriptions is calculating the relative number of uppercase letters at the start of a word.
    
```{python message=FALSE, warning=FALSE}
df = ( df
    .with_columns([
        # Extract the count of lowercase letters starting a word and store it in a new column
        _.Description.str.extract_all(r"[^A-Za-z][a-z]").list.len().alias("lowercase_start_count"),
        # Extract the count of uppercase letters starting a word and store it in a new column
        _.Description.str.extract_all(r"[^A-Za-z][A-Z]").list.len().alias("uppercase_start_count")
    ])
    .with_columns([
        # Create a boolean column indicating if there are no letters starting a word
        ( (_.lowercase_start_count == 0) & (_.uppercase_start_count == 0) ).alias("no_letter_start"),
        # Create a boolean column indicating if there are more lowercase letters starting words than uppercase
        (  _.lowercase_start_count > _.uppercase_start_count ).alias("largely_lowercase_start")
    ])
    .with_columns([
        # Create a final boolean column indicating if the description is anomalous
        ( _.largely_lowercase_start | _.no_letter_start | _.Description.is_null() ).alias("is_anomalous_description")
    ])
    # Drop the intermediate columns used for calculations
    .drop(["lowercase_start_count", "uppercase_start_count", "no_letter_start", "largely_lowercase_start"])
)
```

```{python eval=FALSE, message=FALSE, warning=FALSE}
#  Verify that all descriptions classified as anomalous are so indeed
print(df.filter(_.is_anomalous_description)["Description"].unique().to_list())
```

```{python eval=FALSE, message=FALSE, warning=FALSE}
# Check all the remaining product descriptions
print(df.filter(~_.is_anomalous_description)["Description"].unique().to_list())
```

- This applies to 4193 rows, i.e., $0.4\%$ of the records

```{python message=FALSE, warning=FALSE}
nrows_anomalous_description = df.filter(_.is_anomalous_description).shape[0]
nrows_total = df.shape[0]
(nrows_anomalous_description, nrows_anomalous_description / nrows_total)
```

```{python message=FALSE, warning=FALSE}
# Extract the transaction IDs associated with anomalous descriptions
anomalous_transactions = df.filter(_.is_anomalous_description).select(_.InvoiceNo).with_columns([pl.lit(True).alias("is_anomalous_invoice")])

df = (df
    # Join the main dataframe with the anomalous transactions to mark them
    .join(anomalous_transactions, on="InvoiceNo", how="left")
    # Filter out the rows where the invoice is marked as anomalous
    .filter(_.is_anomalous_invoice.is_null())
    # Drop the columns used for marking and filtering
    .drop(["is_anomalous_description", "is_anomalous_invoice"])
)
```

#### 2.3 Separate Transaction and Product Information

- Here, we separate the transaction and product information

```{python message=FALSE, warning=FALSE}
transactions_df = df.select(_.InvoiceNo, _.StockCode)
products_df = df.select(_.StockCode, _.Description).unique()
```

### 3. Product Description Embeddings

#### 3.1 Create Unique Product Descriptions

```{python message=FALSE, warning=FALSE}
# Group by StockCode and count the occurrences, then sort by count in descending order and get the list of StockCodes
stock_codes = (products_df
    .group_by(_.StockCode)
    .len()
    .sort(_.len, descending=True)
)["StockCode"].to_list()

# Create a pivot table with StockCode as the index and Description as the values
unique_product_descriptions = (products_df
    # Add a row index for each StockCode
    .with_columns( pl.arange(0, pl.len()).over("StockCode").alias("row_idx"))
    .pivot( values="Description", index="row_idx", on="StockCode", aggregate_function="first")
)[stock_codes]
```

```{python message=FALSE, warning=FALSE}
# Display the pivot table
unique_product_descriptions
```

```{python message=FALSE, warning=FALSE}
product_descriptions = products_df.group_by("StockCode").agg(
    pl.col("Description").str.concat(" / ").alias("Description")
)
```

```{python message=FALSE, warning=FALSE}
product_descriptions
```

#### 3.2 Create Description Embeddings


```{python message=FALSE, warning=FALSE}


def cache(func):
    """
    Decorator: Caches the result of func to a pickle file if fname_cache is given.
    The decorated function must accept fname_cache as a keyword argument.
    """
    @wraps(func)
    def wrapper(*args, **kwargs):
        fname_cache = kwargs.get("fname_cache", None)
        
        if fname_cache is not None and os.path.exists(fname_cache):
            print(f"[CACHE] Loading from {fname_cache}")
            with open(fname_cache, "rb") as f:
                return pickle.load(f)
        
        print(f"[CACHE] Computing and caching to {fname_cache}")
        result = func(*args, **kwargs)
        
        if fname_cache is not None:
            with open(fname_cache, "wb") as f:
                pickle.dump(result, f)
            print(f"[CACHE] Saved to {fname_cache}")
        
        return result

    return wrapper
  
@cache
def encode_semantic_embeddings(description, prompt = None):
  """ """
  py_model = st.SentenceTransformer(py_model_name)
  
  prompts = [f"passage: Product: '{t}'" for t in description]
  
  embeddings = py_model.encode(
      sentences = prompts, 
      prompt = prompt,
      show_progress_bar = True,
      use_multiprocessing=True,
      normalize_embeddings = True # note: apparently dot products can be used instead of cosine sim, if normalized
  )

  return embeddings
```

```{python message=FALSE, warning=FALSE}

product_ids = product_descriptions["StockCode"].to_list()
description = product_descriptions["Description"].to_list()

prompts = [f"passage: Product: '{t}'" for t in description]

#description = [
#    "buy", "purchase",
#    "cold", "chilly",
#    "like", "enjoy",
#    "finish", "complete",
#    "meeting", "appointment"
#]
#product_ids = [str(product_id) for product_id in range(0, 10)]

fname_embeddings = "./analysis1_embeddings_semantic.pkl"
norm_embeddings_semantic = encode_semantic_embeddings(description, fname_cache=fname_embeddings)

```

## 4. Product Co-Occurrence Embeddings

- similar (almost the same?) as product2vec: https://github.com/TheoVall-DS/product2vec/ 

```{python message=FALSE, warning=FALSE}
transactions = transactions_df.group_by("InvoiceNo").agg(
    pl.col("StockCode").cast(pl.Utf8).alias("products")
)["products"].to_list()
```

```{python message=FALSE, warning=FALSE}
def resample_transactions(transactions):
    """Resample and shuffle transactions to create a new dataset."""
    
    # Define a lambda function to shuffle a list
    shuffle = lambda lst: random.sample(lst, k=len(lst))

    # Resample the transactions to create a new dataset
    resampled_transactions = shuffle(transactions)  # Resample with replacement
    shuffled_transactions = [shuffle(transaction) for transaction in resampled_transactions]  # Shuffle the sentences

    return shuffled_transactions
```

```{python message=FALSE, warning=FALSE}

#@compute_memory.cache
def encode_transactional_embeddings(product_ids, transactions, embedding_size=100, window=5, min_count=1):
    """ Learn transactional embeddings with Word2Vec on baskets, aggregate for the given product_ids, and return normalized embeddings.
    """
    # Train Word2Vec once (CBOW: sg=0)
    model = Word2Vec(
        transactions,
        vector_size=embedding_size,
        window=window,
        min_count=min_count,
        sg=0,
        workers=4
    )

    # Extract embeddings for requested product_ids
    embeddings = []
    for product_id in product_ids:
        if product_id in model.wv:
            emb = model.wv[product_id]
        else:
            emb = np.full(embedding_size, np.nan)  # fallback if missing
        embeddings.append(emb)

    embeddings = np.array(embeddings)

    # L2-normalize each row
    norms = np.linalg.norm(embeddings, axis=1, keepdims=True)
    norms[norms == 0] = 1  # avoid division by zero
    embeddings /= norms

    return embeddings
```

```{python message=FALSE, warning=FALSE}

@compute_memory.cache
def encode_transactional_embeddings_(product_ids, transactions, embedding_size = 100, window = 5, min_count = 1, n_bootstrap = 1): #fname_cache = None
    """ """

    #if (not fname_cache is None) and os.path.exists(fname_cache):
    #    embeddings = None
    #    with open(fname_cache, "rb") as f:
    #        bootstrap_embeddings, embeddings = pickle.load(f)
    #    return np.array( embeddings )

    # Number of resamples
    bootstrap_embeddings = []

    for _ in tqdm( range(n_bootstrap) ):
        resampled_transactions = resample_transactions(transactions)  # Resample with replacement and re-shuffle the transactions
        model = Word2Vec(transactions, vector_size=embedding_size, window=window, min_count=min_count, sg=0, workers=4)
        bootstrap_embeddings.append({word: model.wv[word] for word in model.wv.index_to_key})
        
    #if (not fname_cache is None):
    #    with open(fname_cache, "wb") as f:
    #        pickle.dump(bootstrap_embeddings, f)

    #bootstrap_embeddings = [ defaultdict(lambda: np.zeros(embedding_size), d) for d in bootstrap_embeddings ]

    #embeddings = []
    #for product_id in tqdm( product_ids ):
    #    product_embeddings = [embedding[product_id] for embedding in bootstrap_embeddings]
    #    aggregated_embedding = np.mean(product_embeddings, axis=0)
    #    embeddings.append( aggregated_embedding )

    #bootstrap_embeddings = [ dict(d) for d in bootstrap_embeddings ]
    #if (not fname_cache is None):
    #    with open(fname_cache, "wb") as f:
    #        pickle.dump((bootstrap_embeddings, embeddings), f)

    return np.array( embeddings )
```

```{python message=FALSE, warning=FALSE}
```

```{python message=FALSE, warning=FALSE}
#%%time
transactional_embedding_size = 100
#transactional_window = 5
transactional_window = np.max([len(x) for x in transactions]) - 1
transactional_min_count = 1
transactional_n_bootstrap = 1

len(transactions)

norm_embeddings_transactional = encode_transactional_embeddings( product_ids, transactions, 
                                                                 embedding_size = transactional_embedding_size, 
                                                                 window = transactional_window,
                                                                 min_count = transactional_min_count)

```

```{python message=FALSE, warning=FALSE}

```


## 5. Similarity Quantification

```{python message=FALSE, warning=FALSE}

sim_semantic = cosine_similarity(norm_embeddings_semantic, norm_embeddings_semantic)
sim_transactional = cosine_similarity(norm_embeddings_transactional, norm_embeddings_transactional)


#sim = (sim - np.min(sim) ) / ( 1 - np.min(sim) )
#sim = np.clip(sim, a_min=0, a_max=1)

```

```{r}

library(reticulate)
library(tidyverse)
library(magrittr)

sim_semantic = py$sim_semantic
sim_transactional = py$sim_transactional

dim(sim_semantic)
dim(sim_transactional)

diag(sim_semantic) <- NA
diag(sim_transactional) <- NA

#n = 4896
#sim_semantic[1:n, 1:n]
#sim_transactional[1:n, 1:n]

vsim_semantic = sim_semantic[lower.tri(sim_semantic)] %>% as.vector()
vsim_transactional = sim_transactional[lower.tri(sim_transactional)] %>% as.vector()

n = 2000
x = vsim_semantic[1:n]
y = vsim_transactional[1:n]
cor(x, y)

data.frame(semantic=x, transactional=y) %>% 
ggplot(aes(semantic, transactional)) + geom_point()

cor( vsim_semantic, vsim_transactional )

```



## 5. Fused Embeddings

```{python message=FALSE, warning=FALSE}

```

```{python message=FALSE, warning=FALSE}
#norm_embeddings_transactional = normalize_embeddings(embeddings_transactional)
#embeddings = np.column_stack((norm_embeddings_semantic, norm_embeddings_transactional))
#embeddings = norm_embeddings_semantic
#embeddings.shape
```

## 6. Similarity Quantification


```{python message=FALSE, warning=FALSE}
sim = cosine_similarity(embeddings, embeddings)
#sim = (sim - np.min(sim) ) / ( 1 - np.min(sim) )
#sim = np.clip(sim, a_min=0, a_max=1)

distance = 1 - sim

from scipy.spatial.distance import squareform
condensed_distance = squareform(distance, checks=False)

from scipy.cluster.hierarchy import linkage, dendrogram

Z = linkage(condensed_distance, method='average')  # or 'average', 'complete', etc.
Z = np.clip(Z, a_min=0, a_max=None)
```

```{python message=FALSE, warning=FALSE}

def compute_homogeneity(indices, similarity_matrix):
    similarity_submatrix = similarity_matrix[np.ix_(indices, indices)]
    homogeneity = np.min( similarity_submatrix )
    return float(homogeneity)

def collect_leaf_indices(node):
    if node.is_leaf():
        return [node.id]
    else:
        return collect_leaf_indices(node.left) + collect_leaf_indices(node.right)

def build_tree(node, labels, similarity_matrix):
    """ 
    """

    if node.is_leaf():
        return [node.id], { # , 
            "name": f"{labels[node.id]}",
            "homogeneity": 0.0  # Leaves have no homogeneity
        }

    else:
        left_leaves, left_branch = build_tree(node.left, labels, similarity_matrix)
        right_leaves, right_branch = build_tree(node.right, labels, similarity_matrix)
        leaves = left_leaves + right_leaves
        
        homogeneity = compute_homogeneity(leaves, similarity_matrix)
        if not left_branch["homogeneity"]:
            left_branch["homogeneity"] = homogeneity

        if not right_branch["homogeneity"]:
            right_branch["homogeneity"] = homogeneity

        return leaves, { 
            "name": f"[{round(100*homogeneity)}]", # "", # or str(node.id)
            "homogeneity": homogeneity,  # Proxy: deeper == more homogeneous
            "children": [ left_branch, right_branch ]
        }


scaled_sim = ( sim - np.min(sim) ) / ( 1 - np.min(np.min(sim)) )

root_node = to_tree(Z)
#x = root_node.right.right.right.left.right
leaves, tree_json = build_tree(root_node, description, sim)
#print("---")
#print(leaves)
#print(tree_json)

with open('tree.json', 'w') as f:
    json.dump(tree_json, f)
```


## Repository

All data and source code are available here: 👉 <https://github.com/plogacev/case_studies/tree/main/product_similarity>
