---
title: "Data Preparation: Panel to Series"
author: "Pavel Logačev"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
engine: knitr
editor: visual
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(data.table)
library(arrow)

cache_dir <- "./_cache"
artifacts_dir <- "./_artifacts"
dir.create(cache_dir, showWarnings = FALSE)
dir.create(artifacts_dir, showWarnings = FALSE)
```

## Summary

This notebook converts raw scanner data from panel format into individual time series suitable for time series modeling. Each product is separated into its own series with daily aggregation of quantity and unit pricing. Series are filtered based on minimum data quality requirements.

## Load and Clean Data

```{r load-data, warning=FALSE, message=FALSE}
df <- fread("./data/scanner_data.csv")

df <- df %>%
  rename(
    product_id = SKU,
    sku_category = SKU_Category,
    date = Date,
    quantity = Quantity,
    sales_amount = Sales_Amount
  ) %>%
  mutate(
    date = as.Date(date, format = "%d/%m/%Y"),
    quantity = as.numeric(quantity),
    sales_amount = as.numeric(sales_amount)
  ) %>%
  filter(quantity > 0, sales_amount > 0) %>%
  mutate(unit_price = sales_amount / quantity)
```

## Aggregate to Daily Level

```{r aggregate-daily}
panel_daily <- df %>%
  group_by(product_id, date) %>%
  summarize(
    quantity = sum(quantity),
    avg_unit_price = sum(sales_amount) / sum(quantity),
    .groups = "drop"
  ) %>%
  arrange(product_id, date)
```

## Fill Missing Dates

For each product, we create a complete daily sequence and fill missing values appropriately.

```{r fill-dates}
fill_series <- function(group_df) {
  date_range <- seq(min(group_df$date), max(group_df$date), by = "day")

  full_df <- data.frame(date = date_range) %>%
    left_join(group_df, by = "date") %>%
    mutate(
      product_id = first(na.omit(product_id)),
      y = replace_na(quantity, 0),
      log_price = log(avg_unit_price)
    )

  # Forward-fill then backward-fill log_price, then use median if still NA
  log_price_median <- median(full_df$log_price, na.rm = TRUE)
  full_df$log_price <- zoo::na.locf(full_df$log_price, na.rm = FALSE)
  full_df$log_price <- zoo::na.locf(full_df$log_price, fromLast = TRUE, na.rm = FALSE)
  full_df$log_price <- replace_na(full_df$log_price, log_price_median)

  # log_y with NA for zeros
  full_df$log_y <- ifelse(full_df$y > 0, log(full_df$y), NA_real_)

  full_df %>% select(product_id, date, y, log_y, log_price)
}

panel_daily_filled <- panel_daily %>%
  group_by(product_id) %>%
  group_split() %>%
  map_dfr(fill_series)
```

## Filter Eligible Series

We apply quality filters to ensure each series has sufficient data for reliable estimation.

```{r filter-series}
min_obs <- 60
min_nonzero <- 10
min_price_unique <- 8

series_diagnostics <- panel_daily_filled %>%
  group_by(product_id) %>%
  summarize(
    n_obs = n(),
    n_nonzero = sum(y > 0),
    n_price_unique = n_distinct(log_price),
    .groups = "drop"
  )

eligible_series <- series_diagnostics %>%
  filter(
    n_obs >= min_obs,
    n_nonzero >= min_nonzero,
    n_price_unique >= min_price_unique
  ) %>%
  mutate(series_id = row_number() - 1L)

panel_daily_final <- panel_daily_filled %>%
  inner_join(eligible_series %>% select(product_id), by = "product_id")

cat("Total series:", nrow(series_diagnostics), "\n")
cat("Eligible series:", nrow(eligible_series), "\n")
```

## Save Artifacts

```{r save-artifacts}
write_parquet(panel_daily_final, file.path(artifacts_dir, "panel_daily.parquet"))
write_parquet(eligible_series, file.path(artifacts_dir, "series_index.parquet"))
```

```{r summary-table}
eligible_series %>%
  summarize(
    n_series = n(),
    median_obs = median(n_obs),
    median_nonzero = median(n_nonzero),
    median_price_unique = median(n_price_unique)
  )
```
