---
title: "Fit Local-Level Model Per Series"
author: "Pavel Logačev"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
engine: knitr
editor: visual
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(TMB)
library(parallel)

artifacts_dir <- "./_artifacts"
```

## Summary

This notebook fits a local-level state-space model with Fourier seasonality to each time series using TMB (Template Model Builder). The model estimates price elasticity while accounting for stochastic trends and weekly seasonal patterns.

## Model Specification

The local-level model with Fourier seasonality is:

**State equation:**
$$\alpha_{t+1} = \alpha_t + \eta_t, \quad \eta_t \sim N(0, \sigma_\eta^2)$$

**Observation equation:**
$$y_t = \alpha_t + \beta \cdot x_t + \sum_{k=1}^{K} \left[\gamma_k \sin\left(\frac{2\pi k t}{7}\right) + \delta_k \cos\left(\frac{2\pi k t}{7}\right)\right] + \varepsilon_t$$

where $\varepsilon_t \sim N(0, \sigma_\varepsilon^2)$, $y_t$ is log quantity, $x_t$ is log price, and $\beta$ is the price elasticity.

## Compile TMB Model

```{r compile-tmb, warning=FALSE, message=FALSE, results='hide'}
compile("local_level_fourier.cpp")
dyn.load(dynlib("local_level_fourier"))
```

## Load Data

```{r load-data}
panel_daily <- read_parquet(file.path(artifacts_dir, "panel_daily.parquet"))
series_index <- read_parquet(file.path(artifacts_dir, "series_index.parquet"))

panel_df <- as.data.frame(panel_daily)
series_index_df <- as.data.frame(series_index)
```

## Fitting Function

```{r fit-function}
n_harmonics <- 3  # Number of Fourier harmonics for weekly seasonality
period <- 7       # Weekly period

fit_series_tmb <- function(product_id_val, series_id_val) {
  series_data <- panel_df %>%
    filter(product_id == product_id_val) %>%
    arrange(date)

  log_y <- series_data$log_y
  log_price <- series_data$log_price
  n <- length(log_y)

  # Find indices of non-NA observations
  obs_idx <- which(!is.na(log_y)) - 1L  # 0-indexed for C++

  # For NA values in log_y, replace with 0 (won't be used in likelihood)
  log_y_filled <- ifelse(is.na(log_y), 0, log_y)

  # Time index (0-indexed)
  t_idx <- 0:(n - 1)

  # TMB data
  data <- list(
    y = log_y_filled,
    x = log_price,
    t_idx = as.integer(t_idx),
    n_harmonics = as.integer(n_harmonics),
    period = as.numeric(period),
    obs_idx = as.integer(obs_idx)
  )

  # Initial parameters
  parameters <- list(
    log_sigma_eta = log(0.1),
    log_sigma_eps = log(0.5),
    beta = -1.0,
    alpha0 = mean(log_y, na.rm = TRUE),
    gamma = rep(0, n_harmonics),
    delta = rep(0, n_harmonics)
  )

  tryCatch({
    # Create TMB object
    obj <- MakeADFun(
      data = data,
      parameters = parameters,
      DLL = "local_level_fourier",
      silent = TRUE
    )

    # Optimize
    opt <- nlminb(
      obj$par,
      obj$fn,
      obj$gr,
      control = list(eval.max = 500, iter.max = 400)
    )

    # Get standard errors via delta method
    sdr <- sdreport(obj)
    sdr_summary <- summary(sdr)

    # Extract beta estimate and SE
    beta_idx <- which(rownames(sdr_summary) == "beta")
    beta_est <- sdr_summary[beta_idx, "Estimate"]
    beta_se <- sdr_summary[beta_idx, "Std. Error"]

    # Extract variance components
    sigma_eta_idx <- which(rownames(sdr_summary) == "sigma_eta")
    sigma_eps_idx <- which(rownames(sdr_summary) == "sigma_eps")

    sigma_eta <- sdr_summary[sigma_eta_idx, "Estimate"]
    sigma_eps <- sdr_summary[sigma_eps_idx, "Estimate"]

    list(
      series_id = series_id_val,
      product_id = product_id_val,
      beta_price = beta_est,
      beta_se = beta_se,
      sigma_eta = sigma_eta,
      sigma_eps = sigma_eps,
      convergence = opt$convergence,
      nll = opt$objective,
      n_obs = length(obs_idx),
      failed = FALSE
    )
  }, error = function(e) {
    list(
      series_id = series_id_val,
      product_id = product_id_val,
      beta_price = NA_real_,
      beta_se = NA_real_,
      sigma_eta = NA_real_,
      sigma_eps = NA_real_,
      convergence = NA_integer_,
      nll = NA_real_,
      n_obs = sum(!is.na(log_y)),
      failed = TRUE
    )
  })
}
```

## Fit All Series

```{r fit-all, warning=FALSE, message=FALSE}
# For demo, limit to first 50 series; set run_full = TRUE for all
run_full <- FALSE
demo_n_series <- 50

if (!run_full) {
  series_to_fit <- series_index_df %>% head(demo_n_series)
} else {
  series_to_fit <- series_index_df
}

cat("Fitting", nrow(series_to_fit), "series...\n")

# Fit in parallel
n_cores <- detectCores() - 1

results_list <- mclapply(
  1:nrow(series_to_fit),
  function(i) {
    fit_series_tmb(
      series_to_fit$product_id[i],
      series_to_fit$series_id[i]
    )
  },
  mc.cores = n_cores
)

# Combine results
fit_results <- bind_rows(results_list)

cat("Completed:", sum(!fit_results$failed), "/", nrow(fit_results), "series\n")
cat("Failed:", sum(fit_results$failed), "series\n")
```

## Results Summary

```{r results-summary}
fit_success <- fit_results %>% filter(!failed)

fit_success %>%
  summarize(
    n_series = n(),
    mean_beta = mean(beta_price, na.rm = TRUE),
    sd_beta = sd(beta_price, na.rm = TRUE),
    median_beta = median(beta_price, na.rm = TRUE),
    mean_se = mean(beta_se, na.rm = TRUE),
    mean_sigma_eta = mean(sigma_eta, na.rm = TRUE),
    mean_sigma_eps = mean(sigma_eps, na.rm = TRUE)
  )
```

```{r beta-histogram, fig.width=7, fig.height=3}
fit_success %>%
  ggplot(aes(x = beta_price)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = median(fit_success$beta_price, na.rm = TRUE),
             linetype = "dashed", color = "red") +
  theme_bw() +
  labs(
    title = "Distribution of Price Elasticity Estimates",
    x = "Price Elasticity (beta)",
    y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Save Results

```{r save-results}
fit_success_df <- fit_results %>% filter(!failed)
fit_failures_df <- fit_results %>% filter(failed)

write_parquet(fit_success_df, file.path(artifacts_dir, "fit_results.parquet"))
write_parquet(fit_failures_df, file.path(artifacts_dir, "fit_failures.parquet"))
```
