---
title: "Parametric Bootstrap Standard Errors"
author: "Pavel Logačev"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
engine: knitr
editor: visual
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(TMB)
library(parallel)

artifacts_dir <- "./_artifacts"
```

## Summary

This notebook computes bootstrap standard errors for the price elasticity estimates from the local-level model. We use parametric bootstrap: simulate new data from the fitted model parameters, refit, and collect the distribution of estimates.

## Load Data and Compiled Model

```{r load-data}
dyn.load(dynlib("local_level_fourier"))

panel_daily <- read_parquet(file.path(artifacts_dir, "panel_daily.parquet"))
fit_results <- read_parquet(file.path(artifacts_dir, "fit_results.parquet"))

panel_df <- as.data.frame(panel_daily)
fit_results_df <- as.data.frame(fit_results)
```

## Bootstrap Configuration

```{r config}
n_boot <- 200
n_harmonics <- 3
period <- 7
max_fail_frac <- 0.5

# For demo, limit series
run_full <- FALSE
demo_n_series <- 25

if (!run_full) {
  fit_results_df <- fit_results_df %>% head(demo_n_series)
}
```

## Parametric Bootstrap Function

The parametric bootstrap simulates new time series from the fitted state-space model:

1. Generate state trajectory: $\alpha_t = \alpha_{t-1} + \eta_t$
2. Generate observations: $y_t^* = \alpha_t + \beta x_t + \text{seasonal}_t + \varepsilon_t$
3. Refit model to simulated data
4. Collect bootstrap distribution of $\hat{\beta}$

```{r bootstrap-function}
bootstrap_series <- function(row) {
  series_id_val <- row$series_id
  product_id_val <- row$product_id

  series_data <- panel_df %>%
    filter(product_id == product_id_val) %>%
    arrange(date)

  log_y <- series_data$log_y
  log_price <- series_data$log_price
  n <- length(log_y)

  # Find observed indices

  obs_mask <- !is.na(log_y)
  obs_idx <- which(obs_mask) - 1L

  if (length(obs_idx) < 20) {
    return(data.frame(
      series_id = series_id_val,
      bootstrap_se = NA_real_,
      bootstrap_ci_lower = NA_real_,
      bootstrap_ci_upper = NA_real_,
      success_rate = 0
    ))
  }

  # Extract fitted parameters
  beta_hat <- row$beta_price
  sigma_eta <- row$sigma_eta
  sigma_eps <- row$sigma_eps
  alpha0 <- mean(log_y, na.rm = TRUE)

  # Get Fourier coefficients (approximate from original fit if not stored)
  # For simplicity, we use zero seasonality in bootstrap simulation
  gamma_hat <- rep(0, n_harmonics)
  delta_hat <- rep(0, n_harmonics)

  t_idx <- 0:(n - 1)

  beta_boot <- numeric(0)
  n_failures <- 0

  for (b in 1:n_boot) {
    tryCatch({
      # Simulate states
      alpha <- numeric(n)
      alpha[1] <- alpha0 + rnorm(1, 0, sigma_eta)
      for (t in 2:n) {
        alpha[t] <- alpha[t - 1] + rnorm(1, 0, sigma_eta)
      }

      # Simulate seasonal component
      seasonal <- rep(0, n)
      for (k in 1:n_harmonics) {
        freq <- 2 * pi * k * t_idx / period
        seasonal <- seasonal + gamma_hat[k] * sin(freq) + delta_hat[k] * cos(freq)
      }

      # Simulate observations (only at observed time points)
      y_star <- rep(NA_real_, n)
      y_star[obs_mask] <- alpha[obs_mask] + beta_hat * log_price[obs_mask] +
                          seasonal[obs_mask] + rnorm(sum(obs_mask), 0, sigma_eps)

      # Prepare data for refitting
      y_star_filled <- ifelse(is.na(y_star), 0, y_star)

      data_boot <- list(
        y = y_star_filled,
        x = log_price,
        t_idx = as.integer(t_idx),
        n_harmonics = as.integer(n_harmonics),
        period = as.numeric(period),
        obs_idx = as.integer(obs_idx)
      )

      parameters <- list(
        log_sigma_eta = log(sigma_eta),
        log_sigma_eps = log(sigma_eps),
        beta = beta_hat,
        alpha0 = alpha0,
        gamma = gamma_hat,
        delta = delta_hat
      )

      obj <- MakeADFun(
        data = data_boot,
        parameters = parameters,
        DLL = "local_level_fourier",
        silent = TRUE
      )

      opt <- nlminb(
        obj$par,
        obj$fn,
        obj$gr,
        control = list(eval.max = 200, iter.max = 150)
      )

      if (opt$convergence == 0) {
        # Extract beta from optimized parameters
        beta_star <- opt$par["beta"]
        if (!is.na(beta_star) && is.finite(beta_star)) {
          beta_boot <- c(beta_boot, beta_star)
        } else {
          n_failures <- n_failures + 1
        }
      } else {
        n_failures <- n_failures + 1
      }
    }, error = function(e) {
      n_failures <<- n_failures + 1
    })
  }

  success_rate <- length(beta_boot) / n_boot

  if (length(beta_boot) < n_boot * (1 - max_fail_frac)) {
    return(data.frame(
      series_id = series_id_val,
      bootstrap_se = NA_real_,
      bootstrap_ci_lower = NA_real_,
      bootstrap_ci_upper = NA_real_,
      success_rate = success_rate
    ))
  }

  data.frame(
    series_id = series_id_val,
    bootstrap_se = sd(beta_boot),
    bootstrap_ci_lower = quantile(beta_boot, 0.025),
    bootstrap_ci_upper = quantile(beta_boot, 0.975),
    success_rate = success_rate
  )
}
```

## Run Bootstrap

```{r run-bootstrap, warning=FALSE, message=FALSE}
cat("Running parametric bootstrap for", nrow(fit_results_df), "series...\n")
cat("Bootstrap samples per series:", n_boot, "\n")

n_cores <- detectCores() - 1

bootstrap_results_list <- mclapply(
  1:nrow(fit_results_df),
  function(i) bootstrap_series(fit_results_df[i, ]),
  mc.cores = n_cores
)

bootstrap_results <- bind_rows(bootstrap_results_list)

cat("Completed bootstrap for", sum(!is.na(bootstrap_results$bootstrap_se)), "series\n")
```

## Results Summary

```{r results-summary}
bootstrap_valid <- bootstrap_results %>%
  filter(!is.na(bootstrap_se))

bootstrap_valid %>%
  summarize(
    n_series = n(),
    mean_bootstrap_se = mean(bootstrap_se, na.rm = TRUE),
    median_bootstrap_se = median(bootstrap_se, na.rm = TRUE),
    mean_success_rate = mean(success_rate, na.rm = TRUE)
  )
```

```{r se-comparison, fig.width=7, fig.height=3}
# Compare bootstrap SE to analytical SE
comparison <- fit_results_df %>%
  inner_join(bootstrap_results, by = "series_id") %>%
  filter(!is.na(bootstrap_se), !is.na(beta_se))

if (nrow(comparison) > 0) {
  comparison %>%
    ggplot(aes(x = beta_se, y = bootstrap_se)) +
    geom_point(alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    theme_bw() +
    labs(
      title = "Analytical vs Bootstrap Standard Errors",
      x = "Analytical SE (TMB)",
      y = "Bootstrap SE"
    ) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

## Save Results

```{r save-results}
write_parquet(bootstrap_results, file.path(artifacts_dir, "bootstrap_results.parquet"))
```
