---
title: "Meta-Analysis of Price Elasticity Estimates"
author: "Pavel Logačev"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
engine: knitr
editor: visual
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)

artifacts_dir <- "./_artifacts"
```

## Summary

This notebook performs meta-analysis on the price elasticity estimates from the local-level models. We use the DerSimonian-Laird method to estimate the overall pooled effect and between-study heterogeneity ($\tau^2$), then compute empirical Bayes shrunken estimates for each series.

## Load Results

```{r load-data}
fit_results <- read_parquet(file.path(artifacts_dir, "fit_results.parquet"))
bootstrap_results <- read_parquet(file.path(artifacts_dir, "bootstrap_results.parquet"))

fit_results_df <- as.data.frame(fit_results)
bootstrap_results_df <- as.data.frame(bootstrap_results)

merged <- fit_results_df %>%
  inner_join(bootstrap_results_df, by = "series_id")

cat("Series with both fit and bootstrap results:", nrow(merged), "\n")
```

## Select Best Standard Error

We prioritize the SE estimate based on reliability:

1. Bootstrap SE if success rate >= 0.6
2. TMB analytical SE otherwise

```{r select-se}
merged <- merged %>%
  mutate(
    se_selected = case_when(
      success_rate >= 0.6 & !is.na(bootstrap_se) ~ bootstrap_se,
      !is.na(beta_se) ~ beta_se,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(se_selected), !is.na(beta_price), se_selected > 0)

cat("Series with valid SE:", nrow(merged), "\n")
```

## DerSimonian-Laird Meta-Analysis

The DerSimonian-Laird method is a random-effects meta-analysis approach that accounts for both within-study variance and between-study heterogeneity.

```{r dersimonian-laird}
dersimonian_laird <- function(beta, se) {
  beta <- as.numeric(beta)
  se <- as.numeric(se)

  # Fixed-effect weights
  w <- 1 / (se^2)
  w_sum <- sum(w)

  # Fixed-effect pooled estimate
  beta_bar <- sum(w * beta) / w_sum

  # Cochran's Q statistic
  Q <- sum(w * (beta - beta_bar)^2)
  k <- length(beta)

  # Estimate tau^2 (between-study variance)
  tau2_num <- Q - (k - 1)
  tau2_den <- w_sum - sum(w^2) / w_sum

  tau2 <- if (tau2_den > 0 && tau2_num > 0) {
    tau2_num / tau2_den
  } else {
    0
  }
  tau2 <- max(0, tau2)

  # Random-effects weights
  w_star <- 1 / (se^2 + tau2)
  w_star_sum <- sum(w_star)

  # Random-effects pooled estimate
  beta_pooled <- sum(w_star * beta) / w_star_sum
  se_pooled <- sqrt(1 / w_star_sum)

  # I^2 statistic (percentage of variance due to heterogeneity)
  I2 <- if (Q > k - 1) {
    100 * (Q - (k - 1)) / Q
  } else {
    0
  }

  list(
    beta_pooled = beta_pooled,
    se_pooled = se_pooled,
    tau2 = tau2,
    tau = sqrt(tau2),
    Q = Q,
    I2 = I2,
    k = k
  )
}
```

## Overall Meta-Analysis

```{r overall-meta}
meta_overall <- dersimonian_laird(merged$beta_price, merged$se_selected)

cat("=== Overall Meta-Analysis Results ===\n")
cat("Number of series:", meta_overall$k, "\n")
cat("Pooled price elasticity:", round(meta_overall$beta_pooled, 4), "\n")
cat("Pooled SE:", round(meta_overall$se_pooled, 4), "\n")
cat("95% CI: [",
    round(meta_overall$beta_pooled - 1.96 * meta_overall$se_pooled, 4), ", ",
    round(meta_overall$beta_pooled + 1.96 * meta_overall$se_pooled, 4), "]\n")
cat("Between-study SD (tau):", round(meta_overall$tau, 4), "\n")
cat("Heterogeneity (I^2):", round(meta_overall$I2, 1), "%\n")
```

## Empirical Bayes Shrinkage

We compute shrunken estimates that pull individual series estimates toward the pooled mean, with the degree of shrinkage depending on each estimate's precision relative to the between-study variance.

$$\hat{\beta}_i^{EB} = \frac{\tau^2}{\sigma_i^2 + \tau^2} \hat{\beta}_i + \frac{\sigma_i^2}{\sigma_i^2 + \tau^2} \bar{\beta}$$

```{r shrinkage}
tau2 <- meta_overall$tau2
beta_pooled <- meta_overall$beta_pooled

merged <- merged %>%
  mutate(
    shrinkage_factor = tau2 / (se_selected^2 + tau2),
    beta_shrunken = shrinkage_factor * beta_price + (1 - shrinkage_factor) * beta_pooled
  )
```

```{r shrinkage-plot, fig.width=7, fig.height=4}
merged %>%
  ggplot(aes(x = beta_price, y = beta_shrunken)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = beta_pooled, linetype = "dotted", color = "red") +
  theme_bw() +
  labs(
    title = "Empirical Bayes Shrinkage of Price Elasticity Estimates",
    subtitle = paste0("Red line: pooled estimate (", round(beta_pooled, 3), ")"),
    x = "Original Estimate",
    y = "Shrunken Estimate"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Stratified Meta-Analysis by Product

```{r meta-by-product}
meta_by_product_list <- merged %>%
  group_by(product_id) %>%
  filter(n() >= 2) %>%
  group_split() %>%
  map(function(df) {
    result <- dersimonian_laird(df$beta_price, df$se_selected)
    data.frame(
      product_id = unique(df$product_id),
      n_series = result$k,
      beta_pooled = result$beta_pooled,
      se_pooled = result$se_pooled,
      tau2 = result$tau2,
      I2 = result$I2
    )
  })

meta_by_product <- bind_rows(meta_by_product_list)

if (nrow(meta_by_product) > 0) {
  meta_by_product %>%
    arrange(beta_pooled) %>%
    head(10)
}
```

## Forest Plot

```{r forest-plot, fig.width=8, fig.height=6}
# Select a subset for visualization
plot_data <- merged %>%
  slice_sample(n = min(30, nrow(merged))) %>%
  arrange(beta_price) %>%
  mutate(
    series_label = paste0("Series ", row_number()),
    ci_lower = beta_price - 1.96 * se_selected,
    ci_upper = beta_price + 1.96 * se_selected
  )

ggplot(plot_data, aes(x = beta_price, y = reorder(series_label, beta_price))) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.2) +
  geom_vline(xintercept = beta_pooled, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50") +
  theme_bw() +
  labs(
    title = "Forest Plot of Price Elasticity Estimates",
    subtitle = paste0("Red dashed line: pooled estimate (", round(beta_pooled, 3), ")"),
    x = "Price Elasticity",
    y = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 8)
  )
```

## Save Results

```{r save-results}
meta_overall_df <- data.frame(
  beta_pooled = meta_overall$beta_pooled,
  se_pooled = meta_overall$se_pooled,
  tau2 = meta_overall$tau2,
  tau = meta_overall$tau,
  I2 = meta_overall$I2,
  k = meta_overall$k
)

shrunken_estimates <- merged %>%
  select(series_id, product_id, beta_price, beta_shrunken, se_selected, shrinkage_factor)

write_parquet(as.data.frame(meta_overall_df), file.path(artifacts_dir, "meta_overall.parquet"))
write_parquet(as.data.frame(meta_by_product), file.path(artifacts_dir, "meta_by_product.parquet"))
write_parquet(as.data.frame(shrunken_estimates), file.path(artifacts_dir, "shrunken_estimates.parquet"))
```

## Interpretation

```{r interpretation}
cat("\n=== Key Findings ===\n\n")
cat("1. Pooled Price Elasticity:", round(beta_pooled, 3), "\n")
cat("   - A 1% increase in price is associated with a",
    round(beta_pooled, 2), "% change in quantity demanded.\n\n")

if (beta_pooled < 0) {
  cat("   - The negative sign indicates demand decreases as price increases (as expected).\n\n")
}

cat("2. Heterogeneity (I^2 =", round(meta_overall$I2, 1), "%):\n")
if (meta_overall$I2 > 75) {
  cat("   - High heterogeneity suggests substantial variation in price sensitivity across products.\n")
} else if (meta_overall$I2 > 50) {
  cat("   - Moderate heterogeneity suggests meaningful variation in price sensitivity across products.\n")
} else {
  cat("   - Low heterogeneity suggests relatively consistent price sensitivity across products.\n")
}

cat("\n3. Between-study SD (tau =", round(meta_overall$tau, 3), "):\n")
cat("   - Individual elasticities vary around the pooled estimate with this standard deviation.\n")
```
